Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-10-02 Manuscript.Rmd
Corrected errors in the app and manuscript tables and figures - growth values were in in2, not cm2.  Corrected version of manuscript included in push.

```{r table-creation}
# State growth
s.g.dat <- state.table.fcn(grow.state, grow.all, 1)

@@

Growth domain comparisons (`r fig_nums("fig_diffgrow")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs. positive estimates, and all three columns of `r fig_nums("fig_diffgrow")` had similar numbers of "significant" comparisons. Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2) or the Above and Below domains (5).  *Thuja plicata* had higher growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3) or the Above and Below domains (3). Other species exhibited no 'significant' findings or displayed a mixture of results.

Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (`r fig_nums("fig_diffmort")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth. There were fewer differences for "Above vs. Below Threshold" comparisons (`r t.ref("tbl_summ_grow_mort")`) than for the comparisons Above or Below the threshold, indicating that mortality may not be as strongly associated with the change in CWD as by the starting value.  *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains. These results occurred in Above (3) or both Above and Below (3) domains. Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.  These species were approximately evenly split between differences being present in the Above (4) or Below (5) domains, but none with 'significant' findings in both the Above and Below domains.

@@

These two are interesting because the lowest mortality domain for PIMO (Above, High) coincides geographically in the southern Sierra with the highest mortality domain for PILA (and for much of the highest-mortality domain for PIPO).  PILA has the highest growth in AH whereas there isn't much of a growth story for PIMA.  PIPO and PICO have great range overlap.  PICO has the highest mortality in Low (NE WA), particularly AL and lowest mortality in BH (Sierras, OR Cascades).  Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for WA and 5 times higher than CA.  PICO growth is lowest in AL.  PIPO has the highest growth in Low.  PIPO in NE WA has much less acreage of pine beetle damage in 2023 than PICO, not really a factor for PIMO.

>>>>

```{r table-creation}
# State growth
grow.state.cm <- cm2.fcn(1, grow.state)
grow.all.cm <- cm2.fcn(1, grow.all)
s.g.dat <- state.table.fcn(grow.state.cm, grow.all.cm, 1)

@@

Growth domain comparisons (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs. positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had similar numbers of "significant" comparisons. Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2) or the Above and Below domains (5).  *Thuja plicata* had higher growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3) or the Above and Below domains (3). Other species exhibited no 'significant' findings or displayed a mixture of results.

Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth. There were fewer differences for "Above vs. Below Threshold" comparisons (`r t.ref("tbl_summ_grow_mort")`) than for the comparisons Above or Below the threshold, indicating that mortality may not be as strongly associated with the change in CWD as by the starting value.  *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains. These results occurred in Above (3) or both Above and Below (3) domains. Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.  These species were approximately evenly split between differences being present in the Above (4) or Below (5) domains, but none with 'significant' findings in both the Above and Below domains.

(Section on geography?  Growth highest for Abies grandis in BL, which coincides with coast range for OR, WA, CA plus Olympics.  CA really only has trees in BL; these probably are the ones that grew the most.  Then, mention some mortality patterns observed below, especially N/NE WA and Sierra Nevadas)

(Thoughts about permutation analysis): The number of significant findings per contrast was unexpected by chance except for Mortality, Above vs. Below.  We examined the likelihood of each of the significant comparisons arising by chance, adjusted by the number of significant findings (120).  Results were considered significant if their permutation p-value was below 0.00041 (0.05/120).  We might need to run 10,000 permutations.

@@

These two are interesting because the lowest mortality domain for PIMO (Above, High) coincides geographically in the southern Sierra with the highest mortality domain for PILA (and for much of the highest-mortality domain for PIPO).  PILA has the highest growth in AH whereas there isn't much of a growth story for PIMA.  PIPO and PICO have great range overlap.  PICO has the highest mortality in Low (NE WA), particularly AL and lowest mortality in BH (Sierras, OR Cascades).  Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for OR and 5 times higher than CA.  PICO growth is lowest in AL.  PIPO has the highest growth in Low.  PIPO in NE WA has much less acreage of pine beetle damage in 2023 than PICO, not really a factor for PIMO.
